---
title: "6372 Stats 2 Project 1"
author: "Chad Kwong"
date: "5/30/2022"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(olsrr)
library(car)
library(ggplot2)
library(class)

#Regression Tree Libraries
library(rsample)     # data splitting 
library(dplyr)       # data wrangling
library(rpart)       # performing regression trees
library(rpart.plot)  # plotting regression trees
library(ipred)       # bagging
library(caret)       # bagging

#loading in data
lifeExpectancyData <- read.csv("/Users/chadkwong/Desktop/Resources/Datasets/Life Expectancy Data.csv")
cleanedData=lifeExpectancyData
#cleanedData$Country = as.factor(Country)
cleanedData$Status = as.factor(cleanedData$Status)
attach(cleanedData)
```

# Objective 1
## Missing Values
### Initial Missing Value Investigation
```{r}
table(is.na(lifeExpectancyData))
#Gathering all na values
#create data frame with 0 rows and 4 columns
naCol <- data.frame(matrix(ncol = 4, nrow = 0))
#provide column names
colnames(naCol) <- c('varName', 'varType', 'varID#', 'naCount')

for(i in 1:ncol(cleanedData)){
  if(length(table(is.na(cleanedData)))==2){
    naCol[i,] = c("dummy","dummy",i,table(is.na(cleanedData[,i]))[2])
  }
}
naCol = na.omit(naCol)
naCol$`varID#` = as.integer(naCol$`varID#`)
naCol$naCount = as.integer(naCol$naCount)
for(i in 1:length(naCol$varName)){
  naCol$varName[i] = colnames(cleanedData)[naCol$`varID#`[i]]
  naCol$varType[i] = typeof(cleanedData[,naCol$`varID#`[i]])
}
naCol

#vis_miss(lifeExpectancyData %>% select_if(~any(is.na(.))))
#gg_miss_upset(lifeExpectancyData)
```
### Approaching NA Values
```{r}
#investigating missing life expectancy values
#Each entry that is missing life expectancy values contain single entries in the dataframe. 
#Therefore I have chosen to remove all rows that have na values for life expectancy
missingLE =  lifeExpectancyData[is.na(lifeExpectancyData$Life.expectancy)==TRUE,]
lifeExpectancyData[is.na(lifeExpectancyData$Life.expectancy)==TRUE,]
#8/10 missing values are for small islands. 2 values are for Monaco and San Marino which are microstates
cleanedData <- cleanedData %>% drop_na(Life.expectancy)
AvgCleanedData = cleanedData

#investigating population
missingPop = cleanedData[is.na(Population)==TRUE,]
popcountrycount = data.frame(table(missingPop$Country))
popcountrycount[order(-popcountrycount$Freq),]
table(popcountrycount$Freq)
missingPopCountries = data.frame("Country" = popcountrycount$Var1)
# Almost all missing populations span across 16 years. 
# This occurs for about 40 countries, so I will not approach fixing these NA values.
# Instead, I will not use this variable as a predictor in my model

```

```{r, include=FALSE}
attach(cleanedData)
```

Hep B
```{r}
#investigating hepatitis B
missingHep = cleanedData[is.na(cleanedData$Hepatitis.B)==TRUE,]
countrycount = data.frame(table(missingHep$Country))
countrycount = countrycount[order(countrycount$Freq),]
countrycount = countrycount[countrycount$Freq!=0,]
countrycount

for(d in 1:length(countrycount$Var1)){
  #Conditional Statement for incomplete countries
  if(countrycount$Freq[d]!=16){
    cleanedData$Hepatitis.B[is.na(cleanedData$Hepatitis.B)==TRUE & Country==countrycount$Var1[d]] = min(na.omit(cleanedData$Hepatitis.B[cleanedData$Country==countrycount$Var1[d]]))
    
  }
  #Conditional Statement for Countries missing all data
  else{
    cleanedData$Hepatitis.B[is.na(cleanedData$Hepatitis.B)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Hepatitis.B))
  }
}

table(is.na(cleanedData$Hepatitis.B))
```

BMI
```{r}
# Approaching BMI
missingBMI = cleanedData[is.na(BMI)==TRUE,]
countrycount = data.frame(table(missingBMI$Country))
countrycount = countrycount[order(countrycount$Freq),]
countrycount = countrycount[countrycount$Freq!=0,]
countrycount

for(d in 1:length(countrycount$Var1)){
  #Conditional Statement for incomplete countries
  if(countrycount$Freq[d]!=16){
    cleanedData$BMI[is.na(cleanedData$BMI)==TRUE & Country==countrycount$Var1[d]] = min(na.omit(cleanedData$BMI[cleanedData$Country==countrycount$Var1[d]]))
    
  }
  #Conditional Statement for Countries missing all data
  else{
    cleanedData$BMI[is.na(cleanedData$BMI)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$BMI))
  }
}

table(is.na(cleanedData$BMI))
```

Diphtheria
```{r}
# Approaching Diphtheria
missingDip = cleanedData[is.na(Diphtheria)==TRUE,]
countrycount = data.frame(table(missingDip$Country))
countrycount = countrycount[order(countrycount$Freq),]
countrycount = countrycount[countrycount$Freq!=0,]
countrycount

for(d in 1:length(countrycount$Var1)){
  #Conditional Statement for incomplete countries
  if(countrycount$Freq[d]!=16){
    cleanedData$Diphtheria[is.na(cleanedData$Diphtheria)==TRUE & Country==countrycount$Var1[d]] = min(na.omit(cleanedData$Diphtheria[cleanedData$Country==countrycount$Var1[d]]))
    
  }
  #Conditional Statement for Countries missing all data
  else{
    cleanedData$Diphtheria[is.na(cleanedData$Diphtheria)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Diphtheria))
  }
}

table(is.na(cleanedData$Diphtheria))
```

Alcohol (setting all missing values to country avg if possible and overall avg where not available). Many missing values were for the most recent year available. After analyzing trends on a rudimentary level, I have chosen to set the missing values to the country average for what data is available.
```{r}
#analyzing random stretch of data
stretch = cleanedData[100:200,c(1,2,7)]
# Approaching Alcohol
missingAlc = cleanedData[is.na(Alcohol)==TRUE,]
countrycount = data.frame(table(missingAlc$Country))
countrycount = countrycount[order(countrycount$Freq),]
countrycount = countrycount[countrycount$Freq!=0,]
countrycount

for(d in 1:length(countrycount$Var1)){
  #Conditional Statement for incomplete countries
  if(countrycount$Freq[d]!=16){
    cleanedData$Alcohol[is.na(cleanedData$Alcohol)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Alcohol[cleanedData$Country==countrycount$Var1[d]]))
    
  }
  #Conditional Statement for Countries missing all data
  else{
    cleanedData$Alcohol[is.na(cleanedData$Alcohol)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Alcohol))
  }
}

table(is.na(cleanedData$Alcohol))
```

Polio seems to be random so I will do random by country when available
```{r}
missingPolio = cleanedData[is.na(cleanedData$Polio)==TRUE,]
countrycount = data.frame(table(missingPolio$Country))
countrycount = countrycount[order(countrycount$Freq),]
countrycount = countrycount[countrycount$Freq!=0,]
countrycount

for(d in 1:length(countrycount$Var1)){
  #Conditional Statement for incomplete countries
  if(countrycount$Freq[d]!=16){
    cleanedData$Polio[is.na(cleanedData$Polio)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Polio[cleanedData$Country==countrycount$Var1[d]]))
    
  }
  #Conditional Statement for Countries missing all data
  else{
    cleanedData$Polio[is.na(cleanedData$Polio)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Polio))
  }
}

table(is.na(cleanedData$Polio))
```

Schooling is missing completely for some countries, so those values will be set to the overall average
```{r}
missingSchooling = cleanedData[is.na(cleanedData$Schooling)==TRUE,]
countrycount = data.frame(table(missingSchooling$Country))
countrycount = countrycount[order(countrycount$Freq),]
countrycount = countrycount[countrycount$Freq!=0,]
countrycount

for(d in 1:length(countrycount$Var1)){
  #Conditional Statement for incomplete countries
  if(countrycount$Freq[d]!=16){
    cleanedData$Schooling[is.na(cleanedData$Schooling)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Schooling[cleanedData$Country==countrycount$Var1[d]]))
    
  }
  #Conditional Statement for Countries missing all data
  else{
    cleanedData$Schooling[is.na(cleanedData$Schooling)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Schooling))
  }
}

table(is.na(cleanedData$Schooling))
```

Income Composition same as schooling
```{r}
missingIncome = cleanedData[is.na(cleanedData$Income.composition.of.resources)==TRUE,]
countrycount = data.frame(table(missingIncome$Country))
countrycount = countrycount[order(countrycount$Freq),]
countrycount = countrycount[countrycount$Freq!=0,]
countrycount

for(d in 1:length(countrycount$Var1)){
  #Conditional Statement for incomplete countries
  if(countrycount$Freq[d]!=16){
    cleanedData$Income.composition.of.resources[is.na(cleanedData$Income.composition.of.resources)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Income.composition.of.resources[cleanedData$Country==countrycount$Var1[d]]))
    
  }
  #Conditional Statement for Countries missing all data
  else{
    cleanedData$Income.composition.of.resources[is.na(cleanedData$Income.composition.of.resources)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Income.composition.of.resources))
  }
}

table(is.na(cleanedData$Income.composition.of.resources))
```

Total Expenditure set to avgs
```{r}
missingTE = cleanedData[is.na(cleanedData$Total.expenditure)==TRUE,]
countrycount = data.frame(table(missingTE$Country))
countrycount = countrycount[order(countrycount$Freq),]
countrycount = countrycount[countrycount$Freq!=0,]
countrycount

for(d in 1:length(countrycount$Var1)){
  #Conditional Statement for incomplete countries
  if(countrycount$Freq[d]!=16){
    cleanedData$Total.expenditure[is.na(cleanedData$Total.expenditure)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Total.expenditure[cleanedData$Country==countrycount$Var1[d]]))
    
  }
  #Conditional Statement for Countries missing all data
  else{
    cleanedData$Total.expenditure[is.na(cleanedData$Total.expenditure)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$Total.expenditure))
  }
}

table(is.na(cleanedData$Total.expenditure))
```

Thinness 5 to 9 missing values will be set to avg for developing countries
```{r}
missingThin59 = cleanedData[is.na(cleanedData$thinness.5.9.years)==TRUE,]
countrycount = data.frame(table(missingThin59$Country))
countrycount = countrycount[order(countrycount$Freq),]
countrycount = countrycount[countrycount$Freq!=0,]
countrycount

cleanedData$thinness.5.9.years[is.na(cleanedData$thinness.5.9.years)==TRUE] = mean(na.omit(cleanedData$thinness.5.9.years[Status=="Developing"]))

table(is.na(cleanedData$thinness.5.9.years))
```

thinness 10 to 19 same as 5 to 9
```{r}
missingThin1019 = cleanedData[is.na(cleanedData$thinness..1.19.years)==TRUE,]
countrycount = data.frame(table(missingThin1019$Country))
countrycount = countrycount[order(countrycount$Freq),]
countrycount = countrycount[countrycount$Freq!=0,]
countrycount

cleanedData$thinness..1.19.years[is.na(cleanedData$thinness..1.19.years)==TRUE] = mean(na.omit(cleanedData$thinness..1.19.years[Status=="Developing"]))

table(is.na(cleanedData$thinness..1.19.years))
```

GDP is missing a lot of values, so missing values will be set to the avg gdp per status.
```{r}
missingGDP = cleanedData[is.na(cleanedData$GDP)==TRUE,]
countrycount = data.frame(table(missingGDP$Country))
countrycount = countrycount[order(countrycount$Freq),]
countrycount = countrycount[countrycount$Freq!=0,]
countrycount

for(d in 1:length(countrycount$Var1)){
  #Conditional Statement for Developing Countries
  if(cleanedData$Status[Country==countrycount$Var1[d]][1]=="Developing"){
    cleanedData$GDP[is.na(cleanedData$GDP)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$GDP[cleanedData$Status=="Developing"]))
    
  }
  #Conditional Statement for Countries missing all data
  else{
    cleanedData$GDP[is.na(cleanedData$GDP)==TRUE & Country==countrycount$Var1[d]] = mean(na.omit(cleanedData$GDP[cleanedData$Status=="Developed"]))
  }
}

table(is.na(cleanedData$GDP))
```

## Initial Data Exploration


```{r, include=FALSE}
attach(cleanedData)
```

We have begun our exploration by setting a seed value and splitting the data 85/15 into training and testing sets respectively.
```{r}
#Initial Data Exploration


set.seed(7777)
index<-sample(1:dim(cleanedData)[1],.85*dim(cleanedData)[1],replace=F)
train<-cleanedData[index,]
test<-cleanedData[-index,]
```


A summary was then conducted on the overall data. 
```{r}

summary(cleanedData)
```

We can see a large difference between the categories for the Status Further exploration was conducted on the status variable.
```{r}
t(aggregate(Life.expectancy~Status,data=cleanedData,summary))
t(aggregate(Life.expectancy~Status,data=cleanedData,sd))
```

From the results above, we can see that there is unequal variance between the two categories, inferring that we should investigate an interactive relationship involving the status variable. Continuing on, correlation plots were recorded for various columns that may be colinear as well as a set of variables that may be useful predictors in predicting Life Expectancy.
```{r}
#Thinness 5 to 9 and Thinness 10 to 18
pairs(cleanedData[,c(19,20)])

#Under Five Deaths and Infant Deaths
pairs(cleanedData[,c(6, 12)])


#everything
#everything = pairs(cleanedData)

# 2 pairs functions for the Selected Variables: Adult Mortality, Under Five Deaths, Percentage Expenditure
# Hepatitis B, Polio, Measles, BMI, HIV Aids, Schooling, Income Composition, Diptheria
pairs(cleanedData[,c(4, 5, 8, 9, 10)])
pairs(cleanedData[,c(4, 11, 12, 13, 15)])
pairs(cleanedData[,c(4, 16, 21, 22)])
```
From the display above, we can see that there is noticeable colinearity between Infant Deaths and Deaths Under Five as well as Thinness 1 to 19 and Thinness 5 to 9. Moving on, I have chosen to find the VIF values of the full model.

```{r}
full.model<-lm(Life.expectancy~.,data=cleanedData[,c(-1)])
alias(lm(Life.expectancy~.,data=cleanedData[,c(-1)]))
vif(full.model)
```

As we can see above, the VIF values for Infant Deaths and Under Five Deaths are extremely high. Based on this and the pairs result above, I have removed the variables infant.deaths and Thinness 10 to 19 years before recalculating the VIFs, I also remove the GDP variable because it seems to be correlated with Percentage Expendature. Income composition and schooling appear to be correlated as well based off of the VIF values. Considering a large correlation constant between schooling and income composition, I have elected to remove 

```{r}
cor(Schooling, Income.composition.of.resources)
```

```{r}
#model with correlated variables found above removed
noCor.model<-lm(Life.expectancy~., data=cleanedData[,c(-1,-6,-17,-19,-23,-24,-25,-26)])
noCor.model<-lm(Life.expectancy~., data = train[,c(-1,-6,-17,-19,-23,-24,-25,-26)])
vif(noCor.model)
```

After removing the Infant death and thinness I have chosen to observe the residuals of the full model with the removed colinear variables
```{r}
# Cooks D Analysis
cooksd <- cooks.distance(noCor.model)
# Plot the Cook's Distance using the traditional 4/n criterion
sample_size <- nrow(cleanedData)
plot(cooksd, pch=".", cex=2, main="Cook's Distance Obs")  # plot cook's distance
abline(h = 4/sample_size, col="red")  # add cutoff line
text(x=1:length(cooksd)+1, y=cooksd, labels=ifelse(cooksd>4/sample_size, names(cooksd),""), col="red")  # add labels

par(mfrow=c(2,2))
plot(noCor.model)
```
Based on the plots above, there is no immediate concern for the residuals. They all appear to be randomly distributed across a normal distribution. The QQ plot also seems to indicate normality. On top of this, a further analysis of the outliers shows that the outlier with a relatively considerable cook's distance is negligible with a value of less than .1.

After importing the olsrr package, a stepwise selection method was performed on the full cleaned model without the Country variable. The reason Country was not included in this models is because the models will produce parameters for each country within the dataset. For the purpose of shortening the interpretation of the model, we have decided to exclude the country variable from our models

```{r}

#performing selection methods on dataset with Country, Infant Deaths, GDP, and Thinness 10 to 19 removed
noCor.selection.model<-lm(Life.expectancy~., data = train[,c(-1,-6,-17,-19)])
ols_step_both_p(noCor.selection.model, prem = 0.05, pent = 0.05, details = FALSE)
ols_step_forward_p(noCor.selection.model, penter=.05, details=FALSE)
ols_step_backward_p(noCor.selection.model)
```
Then further exploration was performed on full models including both Income and Schooling to models with just one of the variables.
```{r}
test.model.1<-lm(Life.expectancy~., data = train[,c(-1,-6,-17,-19)])
test.model.2<-lm(Life.expectancy~., data = train[,c(-1,-6,-17,-19,-21)])
test.model.3<-lm(Life.expectancy~., data = train[,c(-1,-6,-17,-19,-22)])
ols_step_both_p(test.model.1, prem = 0.05, pent = 0.05, details = FALSE)
ols_step_both_p(test.model.2, prem = 0.05, pent = 0.05, details = FALSE)
ols_step_both_p(test.model.1, prem = 0.05, pent = 0.05, details = FALSE)
```
The tables above justify including both Income and schooling in the models

This next chunk performs a second selection process on the variable set obtained in the selection process above to refine our model even further.
```{r}
v2.selection = lm(train$Life.expectancy~., data = train[,c(5, 8,9, 10, 11, 12, 13, 15, 16, 21, 22)])
stepwise = ols_step_both_p(v2.selection, prem = 0.05, pent = 0.05, details = FALSE)
stepwise
#ols_step_backward_p(v2.selection, prem = 0.05, pent = 0.05, details = FALSE)
#ols_step_forward_p(v2.selection, prem = 0.05, pent = 0.05, details = FALSE)

final.simple.model = lm(Life.expectancy~Adult.Mortality+Income.composition.of.resources+Schooling+HIV.AIDS+Polio+BMI+percentage.expenditure+Diphtheria, data = train)
summary(final.simple.model)
confint(final.simple.model)
MSE = mean(summary(final.simple.model)$residuals^2)
MSE
sqrt(MSE)

#Residual Plots
# Cooks D Analysis
cooksd <- cooks.distance(final.simple.model)
# Plot the Cook's Distance using the traditional 4/n criterion
sample_size <- nrow(train)
plot(cooksd, pch=".", cex=2, main="Cook's Distance Obs")  # plot cook's distance
abline(h = 4/sample_size, col="red")  # add cutoff line
text(x=1:length(cooksd)+1, y=cooksd, labels=ifelse(cooksd>4/sample_size, names(cooksd),""), col="red")  # add labels
#plotting residual plots
par(mfrow=c(2,2))
plot(final.simple.model)
```

# Objective 2

## Complex MLR

```{r}
pairs(cleanedData[,c(4, 5, 8, 9, 10, 11, 12, 13, 15, 16, 21, 22)])
```

```{r}
#investigating complex relationships 
par(mfrow=c(1,1))
plot(x = Income.composition.of.resources, y = Schooling)
par(mfrow=c(2,1))
plot(x = Income.composition.of.resources * Schooling, y = Life.expectancy)
plot(x = sqrt(Income.composition.of.resources)*Schooling, y = Life.expectancy)
par(mfrow=c(1,2))
plot(x = Income.composition.of.resources * Schooling, y = Life.expectancy, xlab = "Income Comp")
plot(x = sqrt(Income.composition.of.resources)*Schooling, y = Life.expectancy, xlab="(Income Comp)^(1/2)")

par(mfrow=c(1,2))
plot(x = HIV.AIDS, y = Life.expectancy)
plot(x = sqrt(HIV.AIDS), y = Life.expectancy)

par(mfrow=c(1,2))
plot(x = Adult.Mortality, y = Life.expectancy)
plot(x = sqrt(Adult.Mortality), Life.expectancy)


#Adding in transformations and interactions
cleanedData$rootAdultMortality = sqrt(Adult.Mortality)
cleanedData$rootHIVAIDS = sqrt(HIV.AIDS)
cleanedData$IncomeXSchooling = Income.composition.of.resources * Schooling
cleanedData$RootIncomeXSchooling = sqrt(cleanedData$IncomeXSchooling)
```

Complex Model Selection Process
```{r}
#recreating test and training set
set.seed(42)
trainIndices= sample(1:dim(cleanedData)[1],round(.85 * dim(cleanedData)[1]))
train = cleanedData[trainIndices,]
test = cleanedData[-trainIndices,]

#Performing stepwise selection on dataset including transformations with no interaction terms
v3.selection = lm(train$Life.expectancy~., data = train[,c(5, 8, 9, 10, 11, 12, 13, 15, 16, 21, 22, 23, 24, 25,26)])
v3.selection.method = ols_step_both_p(v3.selection, prem = 0.05, pent = 0.05, details = FALSE)
v3.selection.method
mse = mean(v3.selection.method$model$residuals^2)
sqrt(mse)
#ols_step_backward_p(v3.selection, prem = 0.05, pent = 0.05, details = FALSE)
#ols_step_forward_p(v3.selection, prem = 0.05, pent = 0.05, details = FALSE)
```

```{r}
final.complex.model = v3.selection.method$model
summary(final.complex.model)
MSE = mean(summary(final.complex.model)$residuals^2)
MSE
sqrt(MSE)

# Cooks D Analysis
cooksd <- cooks.distance(final.complex.model)
# Plot the Cook's Distance using the traditional 4/n criterion
sample_size <- nrow(train)
plot(cooksd, pch=".", cex=2, main="Cook's Distance Obs")  # plot cook's distance
abline(h = 4/sample_size, col="red")  # add cutoff line
text(x=1:length(cooksd)+1, y=cooksd, labels=ifelse(cooksd>4/sample_size, names(cooksd),""), col="red")  # add labels
#Plotting residual plots
par(mfrow=c(2,2))
plot(final.complex.model)
```

## Nonparametric Models
### knn Classification Models for Predicting Status
```{r}
#KNN Model for predicting status of a country using variables from v3.selection
set.seed(777)
accs = data.frame(accuracy = numeric(30), k = numeric(30))
for(i in 1:30)
{
  classifications = knn(train[,c(5, 21,22,24,25,26,16,15,23,12)], test[,c(5, 21,22,24,25,26,16,15,23,12)], train$Status, prob = TRUE, k = i, use.all = TRUE)
  table(test$Status,classifications)
  CM = confusionMatrix(table(test$Status,classifications))
  accs$accuracy[i] = CM$overall[1]
  accs$k[i] = i
}
ggplot(accs, aes(x = k, y = accuracy)) + geom_line() #Best K is 3

#KNN Model for predicting status of a country using variables from v3.selection + life expectancy
accs = data.frame(accuracy = numeric(30), k = numeric(30))
for(i in 1:30)
{
  classifications = knn(train[,c(4, 5, 21,22,24,25,26,16,15,23,12)], test[,c(4, 5, 21,22,24,25,26,16,15,23,12)], train$Status, prob = TRUE, k = i, use.all = TRUE)
  table(test$Status,classifications)
  CM = confusionMatrix(table(test$Status,classifications))
  accs$accuracy[i] = CM$overall[1]
  accs$k[i] = i
}
ggplot(accs, aes(x = k, y = accuracy)) + geom_line() #Best K is 7

```

### KNN Regression Models
```{r}
#recreating test and training set seed value = 42
#dropping population column
cleanedData2 = cleanedData[,-18]
cleanedData2$Country=as.factor(cleanedData2$Country)
#Converting Status to binary numeric for inclusion in KNN Regression model: 1 -> Developed, 0 -> Developing
levels(cleanedData2$Status) =  c(1,0)
cleanedData2$Status = as.numeric((cleanedData2$Status))
set.seed(42)
trainIndices= sample(1:dim(cleanedData2)[1],round(.85 * dim(cleanedData2)[1]))
train = cleanedData2[trainIndices,]
test = cleanedData2[-trainIndices,]

#Model 1 regression using full model including status
# extracting x input and y labels 
train_x = train[,c(-1,-4)]
train_x = scale(train_x)[,]
train_y = train[,4]
test_x = test[,c(-1,-4)]
test_x = scale(test_x)[,]
test_y = test[,4]
knnmodel = knnreg(train_x, train_y)
str(knnmodel)
pred_y = predict(knnmodel, data.frame(test_x))

#print(data.frame(test_y, pred_y))

mse = mean((test_y - pred_y)^2)
mae = caret::MAE(test_y, pred_y)
rmse = caret::RMSE(test_y, pred_y)

cat("MSE: ", mse, "MAE: ", mae, " RMSE: ", rmse)

x = 1:length(test_y)
test_y = test_y[order(test_y)]
pred_y = pred_y[order(pred_y)]
plot(x, test_y, col = "red", type = "l", lwd=2,
     main = "Life Expectancy Predictions")
lines(x, pred_y, col = "blue", lwd=2)
legend("topright",  legend = c("Actual Value", "Predicted Value"), 
       fill = c("red", "blue"), col = 2:3,  adj = c(0, 0.6))
grid()
```

```{r}
#Model 1 regression using full model EXCLUDING status
# extracting x input and y labels 
train_x = train[,c(-1,-3,-4)]
train_x = scale(train_x)[,]
train_y = train[,4]
test_x = test[,c(-1,-3,-4)]
test_x = scale(test_x)[,]
test_y = test[,4]
set.seed(42)
knnmodel = knnreg(train_x, train_y)
str(knnmodel)
pred_y = predict(knnmodel, data.frame(test_x))

#print(data.frame(test_y, pred_y))

mse = mean((test_y - pred_y)^2)
mae = caret::MAE(test_y, pred_y)
rmse = caret::RMSE(test_y, pred_y)

cat("MSE: ", mse, "MAE: ", mae, " RMSE: ", rmse)

x = 1:length(test_y)
test_y = test_y[order(test_y)]
pred_y = pred_y[order(pred_y)]
plot(x, test_y, col = "red", type = "l", lwd=2,
     main = "Life Expectancy Predictions")
lines(x, pred_y, col = "blue", lwd=2)
legend("topright",  legend = c("Actual Value", "Predicted Value"), 
       fill = c("red", "blue"), col = 2:3,  adj = c(0, 0.6))
grid()
```

```{r}
#KNN Regression Model Using variables from selection 3
# extracting x input and y labels 
train_x = train[,c(5, 20,21,23,24,25,16,15,22,12)]
train_x = scale(train_x)[,]
train_y = train[,4]
test_x = test[,c(5, 20,21,23,24,25,16,15,22,12)]
test_x = scale(test_x)[,]
test_y = test[,4]
set.seed(42)
knnmodel = knnreg(train_x, train_y)
str(knnmodel)
pred_y = predict(knnmodel, data.frame(test_x))

#print(data.frame(test_y, pred_y))

mse = mean((test_y - pred_y)^2)
mae = caret::MAE(test_y, pred_y)
rmse = caret::RMSE(test_y, pred_y)

cat("MSE: ", mse, "MAE: ", mae, " RMSE: ", rmse)

x = 1:length(test_y)
test_y = test_y[order(test_y)]
pred_y = pred_y[order(pred_y)]
plot(x, test_y[1:100], col = "red", type = "l", lwd=2,
     main = "Life Expectancy Predictions")
lines(x, pred_y[1:100], col = "blue", lwd=2)
#legend("topright",  legend = c("Actual Value", "Predicted Value"), 
       #fill = c("red", "blue"), col = 2:3,  adj = c(0, 0.6))
grid()
```

### Regression Tree Models
full model
```{r}
tree.model <- rpart(
  formula = Life.expectancy ~ .,
  data    = train,
  method  = "anova"
)

#rpart.plot(tree.model)
plotcp(tree.model)
printcp(tree.model)
```

using selection 3 variables
```{r}
tree.model <- rpart(
  formula = Life.expectancy ~ .,
  data    = train[,c(1,4,5, 20,21,23,24,25,16,15,22,12)],
  method  = "anova"
)

#rpart.plot(tree.model)
plotcp(tree.model)
printcp(tree.model)
```

