---
title: "Project 1 - EDA"
author: "Robert Price"
date: "5/30/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 01 - Introduction

For this analysis, we will use the life expectancy dataset collected by the 
World Health Organization (WHO) across multiple years and many countries.
Countries are divided into categories such as "developing" and "developed", and
include metrics collected across a wide array of indicators ranging from adult
mortality to GDP.

The analysis will be completed using the statistical programming language, R.
Third-party packages used in the analysis are listed below. 

#### Load Packages

```{r load_libraries, comment=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(ggthemes)
library(scales)
library(broom)
library(naniar)
library(caret)
library(Metrics)
library(corrplot)
library(mice)
library(imputeTS)
library(e1071)
library(gridExtra)
library(olsrr)
```

## 02 - Summary of Data Set

#### Column Descriptions

- **Country:** country name
- **Year:** year
- **Status:** country status of either "developed or "developing"
- **Life Expectancy:** average life expectancy in years
- **Adult Mortality:** the number of individuals dying between 15 and 60 years 
  old per 1000 population
- **Infant Deaths:** the number of infant deaths per 1000 population
- **Alcohol:** per capita alcohol consumption (in liters of pure alcohol)
- **Percentage Expenditure:** expenditure on health as a percentage of GDP per 
  capita (%)
- **Hepatitis B:** Hepatitis B immunization coverage among 1 year-olds (%)
- **Measles:** number of reported measles cases per 1000 population
- **BMI:** average body mass index entire population
- **Under-Five Deaths:** number of deaths under five years-old per 1000 
  population
- **Polio:** Polio immunization coverage among 1 year-olds (%)
- **Total expenditure:** general government expenditure on health as a 
  percentage of total government expenditure (%)
- **Diphtheria:** Diphtheria tetanus toxoid and pertussis immunization coverage 
  among 1 year-olds (%)
- **HIV/AIDS:** deaths per 1000 live births of HIV/AIDS (0-4 years)
- **GDP:** gross domestic product per capita (in USD)
- **Population:** population of the country
- **Thinness 1-19 Years:** prevalence of thinness among children and adolescents
  ages 1-19 years old (%)
- **Thinness 5-9 Years:** prevalence of thinness among children for ages 5-9
  years old (%)
- **Income Composition of Resources:** human development index in terms of 
  income composition of resources (index ranging from 0 to 1)
- **Schooling:** number of years of schooling

#### Load Dataset

```{r load_dataset}
# input file name to read
filename <- 'Life Expectancy Data.csv'

# read data from CSV
raw_df <- read.csv(filename)
```

#### Rename Columns

```{r rename_columns}
# rename columns using  to clean up
renamed_df <- raw_df %>% 
  rename_with(~ tolower(gsub("[.]+", "_", .x)))

# print header to verify data
head(renamed_df)
```

#### Summary Statistics

From the descriptive statistics below, we can already see some variables that 
could be a concern when modeling. 

```{r}
# simple calculation of descriptive statistics
summary(renamed_df)
```

## 03 - Data Cleaning

#### Visualize Missing Data

```{r missing_data}
# visualize missing observations by column
vis_miss(
  renamed_df %>% select_if(~ any(is.na(.))),
  sort_miss = T
)

# visualize combinations and intersections of missing values
gg_miss_upset(renamed_df)

# create a summary data frame of missing value counts and percentages 
missing_summary_df <- renamed_df %>%
  miss_var_summary() %>%
  mutate(missing = n_miss > 0) %>%
  filter(missing == T)

missing_summary_df

# visualize the percentage of missing variables in the dataset
missing_summary_df %>%
  ggplot(
    aes(x=reorder(variable, pct_miss), 
        y=pct_miss, color=missing, fill=missing)) +
    geom_bar(stat='identity', position='dodge', width=0) + 
    geom_point() +
    coord_flip() +
    scale_colour_manual(values = c('#c0392b', '#8E9796')) +
    scale_fill_manual(values = c('#c0392b', '#8E9796')) +
    labs(title='Missing Values by Feature', x='', y='Percent Missing (%)') +
    theme_minimal() +
    theme(legend.position='none')
```

**Columns with Missing Values**
- **Population:** drop column due to high number of missing values (22.2%) and 
  bias in which countries are missing records
  
- **Life Expectancy:** life expectancy is the response variable, and there is 
  small number of missing values (0.34%), so I will drop these missing records
  from analysis


## Visualizations

#### Visualize Population Over Time

```{r}
# plot total population over time
renamed_df %>% 
  mutate(population = ifelse(is.na(population), 0, population)) %>%
  ggplot(aes(x=year, y=population)) + 
  geom_col(fill='#2c3e50') + 
	scale_y_continuous(labels = label_number(suffix = ' B', scale = 1e-9)) + 
  labs(
    title='Total World Population by Year',
    x='', y='Population in Billions')

# list countries missing population values
renamed_df %>%   
  mutate(population = ifelse(is.na(population), 0, population)) %>%
  group_by(country) %>%
  summarize(count_na=sum(!population > 0)) %>%
  filter(count_na > 0)
```

#### Initial Data Cleaning

```{r}
# count the number of records where life expectancy is missing
count_missing_le <- renamed_df %>% 
  filter(is.na(life_expectancy)) %>% count()
paste0('Number of records missing the response variable, life expectancy: ', 
       count_missing_le)

# drop rows where life expectancy is not defined
drop_le_records_df <- renamed_df %>% drop_na(life_expectancy)

# drop population variable from dataset
drop_population_df <- 
  drop_le_records_df[, !(names(drop_le_records_df) %in% c('population'))]
```

#### Factorize Categorical Variables

```{r}
# define function for factorizing a list of columns
factorize_columns = function (df, cols) {
  for (col in columns_to_factor) {
    df[, col] <- as.factor(df[, col])
  } 
  return (df)
}

# list of columns to factorize
columns_to_factor <- c('country', 'status')

# convert columns to factors
factorized_df <- factorize_columns(drop_population_df, columns_to_factor)
```


#### Missing Value Imputation

```{r, echo=FALSE}
# drop records for testing imputation
imputation_copy_df <- factorized_df %>% drop_na()

# set constants to control calculation
n_miss <- 80
interations <- 20
imputation_methods <- c('Mean', 'Median', 'Group Means') # c('Mean', 'Median', 'MICE')

imputation_results <- setNames(
  data.frame(matrix(ncol=3, nrow=0)), c('Method', 'X', 'RMSE'))

# loop over methods and create scoring metrics
for (method in imputation_methods) {
  if (method == 'Mean') {
    for (i in 1:interations) {

      # create a test of the dataframe to test imputation methods
      imputation_df <- imputation_copy_df
    
      # randomly set null values in dataset
      imputation_df[sample(1:nrow(imputation_df), n_miss), 'hepatitis_b'] <- NA
    
      # mean imputation
      imputer <- imputation_df %>% 
        group_by(country)
        
      # imputer <- Hmisc::impute(imputation_df$hepatitis_b, mean)
    
      # calculate actual and predicted
      actuals <- imputation_copy_df$hepatitis_b[is.na(imputation_df$hepatitis_b)]
      predicted <- rep(mean(imputation_df$hepatitis_b, na.rm=T), length(actuals))
    
      # create new row for dataframe
      new_row <- c(method, i, rmse(actuals, predicted))
      imputation_results[nrow(imputation_results) + 1, ] <- new_row
      
    }
    
  }
  
  if (method == 'Median') {
    for (i in 1:interations) {

      # create a test of the dataframe to test imputation methods
      imputation_df <- imputation_copy_df
    
      # randomly set null values in dataset
      imputation_df[sample(1:nrow(imputation_df), n_miss), 'hepatitis_b'] <- NA
    
      # mean imputation
      imputer <- Hmisc::impute(imputation_df$hepatitis_b, median)
    
      # calculate actual and predicted
      actuals <- imputation_copy_df$hepatitis_b[is.na(imputation_df$hepatitis_b)]
      predicted <- rep(mean(imputation_df$hepatitis_b, na.rm=T), length(actuals))
    
      # create new row for dataframe
      new_row <- c(method, i, rmse(actuals, predicted))
      imputation_results[nrow(imputation_results) + 1, ] <- new_row
      
      }
  }
    
  if (method == 'MICE') {
    for (i in 1:interations) {

      # create a test of the dataframe to test imputation methods
      imputation_df <- imputation_copy_df
    
      # randomly set null values in dataset
      imputation_df[sample(1:nrow(imputation_df), n_miss), 'hepatitis_b'] <- NA
    
      # mean imputation
      mice_model <- mice(imputation_df[, !names(imputation_df) %in% 'life_expectancy'], method="rf")
      mice_output <- complete(mice_model)
      
      actuals <- imputation_copy_df$hepatitis_b[is.na(imputation_df$hepatitis_b)]
      predicted <- mice_output[is.na(imputation_df$hepatitis_b), 'hepatitis_b']

      # create new row for dataframe
      new_row <- c(method, i, rmse(actuals, predicted))
      imputation_results[nrow(imputation_results) + 1, ] <- new_row
      
    }
  }
  
}

```


```{r}

imputation_results$X <- as.integer(imputation_results$X)
imputation_results$RMSE <- as.numeric(imputation_results$RMSE)


imputation_results %>% ggplot(aes(x=X, y=RMSE, color=Method)) + 
  geom_point(shape=1) +
  expand_limits(x = 0, y = 0) + 
  labs(title='Comparison of Imputation Methods', x='')

imputation_results %>%
  group_by(Method) %>%
  summarise(Avg.RMSE=mean(RMSE))
```

#### Fill Missing Values

```{r}
# impute missing values using function
clean_df <- factorized_df %>% 
  group_by(country) %>% 
  mutate_if(
    is.numeric, 
    function(x) ifelse(is.na(x), mean(x, na.rm = T), x)) %>%
  ungroup() %>%
  na_mean()

# plot the distribution of Hepatitis B on a histogram
clean_df %>% ggplot(aes(y=hepatitis_b)) +
  geom_histogram(fill='#2c3e50') + 
  labs(
    title='Distribution of Hepatitis B Vaccinations',
    x='Count', y='Vaccination [%]')

# plot box plots of Hepatitis B by Development Status
factorized_df %>% ggplot(aes(x=status, y=hepatitis_b)) +
  geom_boxplot()

# plot box plots from clean dataframe
clean_df %>% ggplot(aes(x=status, y=hepatitis_b)) +
  geom_boxplot()

# verify that all records have been filled
clean_df %>% miss_var_summary() %>% filter(n_miss > 0)
```

### Boxplots

```{r}
# define negate operator %!in%
`%!in%` <- Negate(`%in%`)

# define columns to skip plotting
columns_to_skip <- c('country', 'status', 'year')

# loop over dataframe and plot box plots 
for(i in names(drop_population_df)) {
  if (i %!in% columns_to_skip) {
      plt <- drop_population_df %>% ggplot(aes_string(x=i)) +
        geom_boxplot(aes(fill=status)) + 
        labs(
          title=paste0('Histogram of "', i, '"'),
          )
      print(plt)
  }
}
```

#### Correlation

```{r}
# define columns to skip correlation
columns_to_skip <- c('year', 'country', 'status')

# create dataframe to prepare for correlation analysis
corr_df <- clean_df[, !(names(clean_df) %in% columns_to_skip)]

# create a function to flatten the correlation matrix
flatten_cor_matix <- function(cor_mat, p_mat) {
  ut <- upper.tri(cor_mat)
  df <- data.frame(
    row = rownames(cor_mat)[row(cor_mat)[ut]],
    column = rownames(cor_mat)[col(cor_mat)[ut]],
    cor  =(cor_mat)[ut],
    p = p_mat[ut]
  )
  return(df)
}

# create flattened matrix with p-values
res2 <- Hmisc::rcorr(as.matrix(corr_df))
flat_cor_matrix <- flatten_cor_matix(res2$r, res2$P)
flat_cor_matrix

# filter for specific variables
flat_cor_matrix %>%
  filter(row == 'percentage_expenditure')

# plot correlation matrix
corrplot(
  cor(corr_df),
  method = 'circle',
  type = 'lower',
  hc.order = T,
  order = 'alphabet',
  title = 'Correlation of Continuous Variables',
  number.cex = 0.5,
  tl.cex = 0.6,
  tl.col = 'black'
)


# plot specific variables on x-y scatter plot
r <- round(
  cor(clean_df$thinness_1_19_years, clean_df$thinness_5_9_years), 2)
clean_df %>% ggplot(aes(x=thinness_1_19_years, y=thinness_5_9_years)) +
  geom_point(shape=1, color='#2980b9') + 
  labs(title='Variable Scatter Plot', subtitle=paste0('Correlation Coefficient (r): ', r))

# plot specific variables on x-y scatter plot
# clean_df %>% ggplot(aes(x=percentage_expenditure, y=gdp)) +
#   geom_point(shape=1, color='#2980b9')

# plot specific variables on x-y scatter plot
# clean_df %>% ggplot(aes(x=life_expectancy, y=adult_mortality)) +
#   geom_point(shape=1, color='#c0392b')

# plot specific variables on x-y scatter plot
r <- round(
  cor(clean_df$under_five_deaths, clean_df$infant_deaths), 3)
clean_df %>% ggplot(aes(x=under_five_deaths, y=infant_deaths)) +
  geom_point(shape=1, color='#2980b9') + 
  labs(title='Variable Scatter Plot', subtitle=paste0('Correlation Coefficient (r): ', r))

# plot specific variables on x-y scatter plot
r <- round(
  cor(clean_df$schooling, clean_df$income_composition_of_resources), 2)
clean_df %>% ggplot(aes(x=schooling, y=income_composition_of_resources)) +
  geom_point(shape=1, color='#2980b9') + 
  labs(title='Variable Scatter Plot', subtitle=paste0('Correlation Coefficient (r): ', r))
```

#### Drop Highly Correlated Variables

```{r}
# define columns to skip correlation
correlated_columns_to_drop <- c('gdp', 'infant_deaths', 'thinness_10_19_years')

# create dataframe to prepare for correlation analysis
no_cor_df <- clean_df[, !(names(clean_df) %in% correlated_columns_to_drop)]
```

```{r}
summary(no_cor_df)
```


## 06 - Feature Engineering & Model 


#### Split Data into Traning and Testing Sets

```{r}
# add ID column to dataset
model_ready_df <- rowid_to_column(no_cor_df, 'id')

#' Function that splits a dataframe into a training set and test set based 
#' on a specified split proportion
#'
#' Args
#'  df (data.frame): input dataframe to be split
#'  split_perc (int): percentage of data assigned to training set 
#'  
#' Returns
#' (list) a list of dataframes: one training and one testing dataset
train_test_split = function(df, split_perc){
  
  train_ind = sample(1:dim(df)[1], round(split_perc * dim(df)[1]))
  train = df[train_ind,]
  test = df[-train_ind,]
  
  return(list(train=train, test=test))
}

# split data into training and testing sets
train_test <- train_test_split(model_ready_df, 0.85)

# assign new dataframes
train_df <- train_test$train
test_df <- train_test$test

# print number of observations in each set
paste0('Number of rows in training set: ', nrow(train_df))
paste0('Number of rows in testing set: ', nrow(test_df))
```

#### Define Model Interpretation Metrics

```{r}
# define a simple metric to measure model complexity using the number of coefficients
get_complexity = function(model) {
  length(coef(model)) - 1
}

# define response name for plots
res_variable <- 'life_expectancy'
res_variable_txt <- 'Life Expectancy'
```

#### Fit and Predict Intercept Model

```{r}
# define simple models to run
simple_models <- c(
  # 'Intercept-Only', 
  'All Variables'
  # 'Intuition', 
  # 'Forward Selection',
  # 'Backward Selection',
  # 'Step-Wise Selection'
  )

# create a dataframe to store model results
model_results <- setNames(
  data.frame(
    matrix(ncol=6, nrow=0)), 
    c('Model', 'TrainRMSE', 'TestRMSE', 'AdjR2', 'R2', 'Predictors'))

# loop over simple models 
for (model in simple_models) {

  # set initial value for model fit
  fit <- NULL
  
  # plot additional plots for step-wise
  is_stepwise <- F
  
  # fit intercept-only model as starting point for feature selection 
  intercept_only <- lm(life_expectancy ~ 1, data=train_df)
  
  # set all variables model as starting point for feature selection
  all <- lm(life_expectancy ~ . - id - country, data=train_df)
  
  # fit the intercept-only model to the training dataset
  if (model == 'Intercept-Only') {
    fit = intercept_only
  }
  
  # fit all variables to the model
  if (model == 'All Variables') {
    fit = all
  }
  
  # fit an intuition model 
  if (model == 'Intuition') {
    fit = lm(life_expectancy ~ 
         adult_mortality +
         income_composition_of_resources +
         schooling + 
         hiv_aids + 
        percentage_expenditure,
       data=train_df)
  } 
  
  # fit a forward selection model
  if (model == 'Forward Selection') {
    is_stepwise = T
    ols_step_forward_p(all, penter=0.05, details=F)
    fit = lm(life_expectancy ~ 
        adult_mortality +
        income_composition_of_resources +
        schooling + 
        hiv_aids + 
        diphtheria + 
        bmi + 
        status + 
        polio + 
        percentage_expenditure + 
        measles, 
      data=train_df)
  }
  
  # fit a backward step-wise regression
  if (model == 'Backward Selection') {
    is_stepwise = T
    ols_step_backward_p(all, prem=0.05, details=F)
    fit = lm(life_expectancy ~ . -id - country - thinness_5_9_years 
             - alcohol - year - hepatitis_b - under_five_deaths,
      data=train_df)
  }
  
  # fit a step-wise regression
  if (model == 'Step-Wise Selection') {
    is_stepwise = T
    ols_step_both_p(all, penter = 0.05, prem = 0.05, details=F)
    fit = lm(life_expectancy ~ 
      adult_mortality +
      income_composition_of_resources +
      schooling + 
      hiv_aids + 
      diphtheria + 
      bmi + 
      status + 
      polio + 
      percentage_expenditure + 
      measles,
    data=train_df)
  }
  
  if (!is.null(fit)) {
    # calculate predictions from the training and testing set
    train_preds <- predict(fit)
    test_preds <- predict(fit, newdata = test_df)
    
    # calculate RMSE for training and testing sets
    train_rmse <- rmse(train_df$life_expectancy, train_preds)
    test_rmse <- rmse(test_df$life_expectancy, test_preds)
    
    # calculate R2 and adjusted R2 metrics
    adj_r2 <- summary(fit)$adj.r.squared
    r2 <- summary(fit)$r.squared
    
    if (is_stepwise) {
      predictor_vars <- paste(k$predictors, collapse=', ')
    } else
      predictor_vars <- ''
    
    # add model results to dataframe
    new_row <- c(
      model, 
      round(train_rmse, 4), 
      round(test_rmse, 4), 
      round(adj_r2, 4), 
      round(r2, 4),
      predictor_vars)
		model_results[nrow(model_results) + 1, ] <- new_row
    

    # set a plot with size 2x2 and plot the diagnostics
    # par(mfrow=c(2,2))
    plot(fit)
    
    
    # join prediction results with ID
    results_df <- data.frame(
        id = test_df$id,
        life_expectancy = test_df$life_expectancy,
        prediction = test_preds
      ) %>%
      mutate(residual = life_expectancy - prediction)
    
    # create a scatter plot of residuals
    p1 <- results_df %>% ggplot(aes(x=id, y=residual)) +
      geom_point(color='#c0392b', shape=1, alpha=0.8) +
      geom_hline(yintercept=0) +
      labs(
        title='Residuals Plot',
        x='Fitted Values',
        y='Residual'
      ) + 
      theme(
        plot.title = element_text(size=10, face='bold'),
        axis.title = element_text(size=9)
      )

    # create a histogram of residuals
    p2 <- results_df %>% ggplot(aes(x=residual)) +
      geom_histogram(aes(y=..density..), fill='#c0392b', color='white', bins=17) +
      labs(
        title='Histogram of Residuals',
        x='Residuals',
        y=NULL
      ) + 
      theme(
        plot.title = element_text(size=10, face='bold'),
        axis.title = element_text(size=9)
      )

    # scatter plot of predictions and residuals
    p3 <- results_df %>% ggplot(aes(x=prediction, y=life_expectancy)) +
      geom_point(color='#0C6291', shape=1, alpha=0.8) +
      geom_abline(intercept=0, slope=1) +
      labs(
        title='Observed vs Predicted',
        x='Predicted',
        y='Observed (Test Set)'
      ) + 
      theme(
        plot.title = element_text(size=10, face='bold'),
        axis.title = element_text(size=9)
      )
    
    g <- grid.arrange(
      p1, p2, p3,
      nrow = 1,
      top = paste0('Diagnostics Plots for ', model, ' Model')
    )
    
    g
        
  }
  

}

# convert model result data types
model_results$TrainRMSE <- as.numeric(model_results$TestRMSE)
model_results$TestRMSE <- as.numeric(model_results$TestRMSE)
model_results$AdjR2 <- as.numeric(model_results$AdjR2)
model_results$R2 <- as.numeric(model_results$R2)
```



```{r}
model_results
```

```{r}

reduced_model <- lm(life_expectancy ~ 
  adult_mortality +
  income_composition_of_resources +
  schooling + 
  hiv_aids + 
  diphtheria + 
  bmi + 
  status + 
  polio + 
  percentage_expenditure + 
  measles,
data=train_df)

ols_step_forward_aic(reduced_model, details=F)

```
